Investigation on Parallel Training of Deep Neural Network using Model Averaging

### 2015-06-24 paper.tex
adding more text and experiment results

% Title.
% ------
\title{Parallel Training of Deep Neural Network using Model Averaging and Butterfly Mixing}
%
% Single address.
% ---------------
\name{Hang Su$^{1,2}$, Haoyu Chen$^1$, Haihua Xu$^3$}
\address{$^1$ International Computer Science Institute, Berkeley, California, US \\
$^2$ Dept. of Electrical Engineering \& Computer Science, University of California, Berkeley, CA, USA \\
$^3$ Nanyang Technological University, Singapore \\
{\small \tt \{suhang3240@gmail.com\}}
}
%
% For example:
% ------------
%\address{School\\
%	Department\\
%	Address}
%
% Two addresses (uncomment and modify for two-address case).
% ----------------------------------------------------------
%\twoauthors
%  {A. Author-one, B. Author-two\sthanks{Thanks to XYZ agency for funding.}}
%	{School A-B\\
%	Department A-B\\
%	Address A-B}
%  {C. Author-three, D. Author-four\sthanks{The fourth author performed the work
%	while at ...}}
%	{School C-D\\
%	Department C-D\\
%	Address C-D}
%
\begin{document}
%\ninept
%
\maketitle
%
\begin{abstract}
In this paper we apply model averaging and butterfly mixing to parallel training of deep neural network (DNN). 
Parallelization is done in a model averaging manner. Data is partitioned and distributed to different nodes for 
local model updates, and model averaging (reduce) is done every few minibatches.

We use multiple GPUs for data parallelization, and Message Passing Interface (MPI) for communication between nodes. 
We investigate the effectiveness of Natrual Gradient for Stochasitc Gradient Descent (NG-SGD) in the model averaging
framework, and compare two learning rate schedule strategies for model update. We show that combination of 
NG-SGD and exponential learning rate schedule can lead to better convergence. On the 300h Swithboard dataset, a 10 
times speed up is achieved using 16 gpus with little decoding accuracy loss -- a 0.2\% degradation in word error rate 
(WER). We also introduce a butterfly mixing strategy that can further save communication costs in model averaging framework.
\end{abstract}
%
\begin{keywords}
Parallel training, model averaging, deep neural network, butterfly mixing
\end{keywords}
%
\section{Introduction}
\label{sec:intro}
Deep Neural Networks (DNN) has shown its effeciveness in several machine learning tasks, espencially in speech
recognition. The large model size and massive training examples make DNN a powerful model for classification. However,
these two factors also slow down the training procedure of DNNs.

@@

Our approach is mainly based on the NG-SGD with model averaging. The main differences here in this work are:
1. We utilize multiple GPUs in neural networks training via MPI, which allows us to perform model averaging more frequently and efficiently;
2. We introduce a butterfly-mixing strategy for model averaging that can further speed up the training.

In addition, this work simplifies the model training procedure in the following aspects:
1. It does not require a warm-up phase where only single thread can be used for model update \cite{seide20141,povey2014parallel};
2. It does not need to tune minibatch size to utilize parallelization power;
3. It does not use momentum or Adagrad\cite{duchi2011adaptive} for gradient estimation.

@@

\section{Data parallelization and Model Averging}
SGD is a popular method for DNN training. Even though neural network training objectives are usually non-convex, 
mini-batch SGD has been shown to be effective for optimizing the objective\cite{seide2011conversational}. 
Roughly speaking, a bigger minibatch size gives a better estimate of the gradient, resulting in a better the converge rate. 
Thus, a straight forward idea for parallellization would be distributing the gradient computation to different computing
nodes. In each step, gradients of minibatches on different nodes are reduced to a single node, averaged and then used to 
update models in each node. This method can compute the gradient accurately, but it requires heavy communication between nodes.
Also, it is shown that increasing minibatch size does not always benefit model training\cite{}, especially in early stage 
of model training.

@@

\section{Model Averaging Strategy}
\subsection{All-reduce}
All-reduce strategy collects the weights from all the nodes in the network. The communication is bounded by 
the bandwidth. Fig~\ref{fig:allreduce} is an example of all-reduce with 4 nodes. One node is responsible for
collecting all the model parameters. It will average the weights and then broadcast it to all the nodes. 
This approach compute the average weight based on all the information in computing nodes. Thus, we expect 
the converge speed to be the fastest.
\begin{figure}[htb]
  \centering
  \includegraphics[width=0.5\textwidth]{allreduce.png}
  \caption{All-reduce network}
  \label{fig:allreduce}
\end{figure}

@@

\section{Experimental Results}
\subsection{Setup}

Reduce frequency

\subsection{Learing Rate Schedule}
Why do we need a large learning rate?

\subsection{Initialization Matters}
Compare random init and dbn init, single and paralle
\begin{table}
  \centering
  \begin{tabular}{c|c|c}
    \hline
     Nodes          &  1    & 16 \\
    \hline
    random init     & --    & --  \\
    \hline
    rbm pretraining & --    & --\\
    \hline
  \end{tabular}
  \caption{Comparing rbm pretraining with random initialization in model averaging setup }
  \label{tab:init}
\end{table}

\subsection{Online NG-SGD Matters}
\begin{table}
  \centering
  \begin{tabular}{c|c|c}
    \hline
     Nodes  &  1    & 16 \\
    \hline
    SGD     & --    & --  \\
    \hline
    NG-SGD  & --    & --\\
    \hline
  \end{tabular}
  \caption{Comparing NG-SGD and naive SGD}
  \label{tab:ngsgd}
\end{table}

@@

\begin{table}
  \centering
  \begin{tabular}{c|c|c|c|c|c}
    \hline
     Nodes  & 1    & 2    & 4    & 8    & 16 \\
    \hline
Allreduce &    14.9 & -- & -- & 15.1 & 15.1 \\
    \hline
Butterfly &    14.9 & -- & -- & -- & --\\
    \hline
  \end{tabular}
  \caption{Comparison of WERs using different number of GPUs }
  \label{tab:wer}
\end{table}

@@

CUDA aware MPI, learning rate schedule.

@@

>>>>

% Title.
% ------
\title{Investigation on Parallel Training of Deep Neural Network using Model Averaging}
%
% Single address.
% ---------------
\name{Hang Su$^{1,2}$, Haoyu Chen$^1$, Haihua Xu$^3$}
\address{$^1$ International Computer Science Institute, Berkeley, California, US \\
$^2$ Dept. of Electrical Engineering \& Computer Science, University of California, Berkeley, CA, USA \\
$^3$ Nanyang Technological University, Singapore \\
{\small \tt \{suhang3240@gmail.com\}}
}
%
% For example:
% ------------
%\address{School\\
%	Department\\
%	Address}
%
% Two addresses (uncomment and modify for two-address case).
% ----------------------------------------------------------
%\twoauthors
%  {A. Author-one, B. Author-two\sthanks{Thanks to XYZ agency for funding.}}
%	{School A-B\\
%	Department A-B\\
%	Address A-B}
%  {C. Author-three, D. Author-four\sthanks{The fourth author performed the work
%	while at ...}}
%	{School C-D\\
%	Department C-D\\
%	Address C-D}
%
\begin{document}
%\ninept
%
\maketitle
%
\begin{abstract}
In this paper we apply model averaging to parallel training of deep neural network (DNN). 
Parallelization is done in a model averaging manner. Data is partitioned and distributed to different nodes for 
local model updates, and model averaging across nodes is done every few minibatches.

We use multiple GPUs for data parallelization, and Message Passing Interface (MPI) for communication between nodes. 
We investigate the effectiveness of Natrual Gradient for Stochasitc Gradient Descent (NG-SGD) in the model averaging
framework, and compare two learning rate schedule strategies for model update. It is shown that NG-SGD leads to better
convergence in model averaging framework. On the 300h Swithboard dataset, a 10 times speedup is achieved using 16 GPUs
and 17 times speedup using 32 GPUs with a small decoding accuracy loss -- 0.4\% degradation in word error rate 
(WER) respectively. We also introduce a butterfly mixing strategy that can further save communication costs in model 
averaging framework.
\end{abstract}
%
\begin{keywords}
Parallel training, model averaging, deep neural network, butterfly mixing
\end{keywords}
%
\section{Introduction}
\label{sec:intro}
Deep Neural Networks (DNN) has shown its effeciveness in several machine learning tasks, espencially in speech
recognition. The large model size and massive training examples make DNN a powerful model for classification. However,
these two factors also slow down the training procedure of DNNs.

@@

Our approach is mainly based on the NG-SGD with model averaging. We utilize multiple GPUs in neural networks training via 
MPI, which allows us to perform model averaging more frequently and efficiently. Currently, this work simplifies the model 
training procedure in the following aspects:
1. It does not add a warm-up phase where only single thread is used for model update \cite{seide20141,povey2014parallel};
2. It does not tune minibatch size to utilize parallelization power;

@@

\section{Data parallelization and Model Averging}
SGD is a popular method for DNN training. Even though neural network training objectives are usually non-convex, 
mini-batch SGD has been shown to be effective for optimizing the objective\cite{seide2011conversational}. 
Roughly speaking, a bigger minibatch size gives a better estimate of the gradient, resulting in a better the converge rate. 
Thus, a straight forward idea for parallellization would be distributing the gradient computation to different computing
nodes. In each step, gradients of minibatches on different nodes are reduced to a single node, averaged and then used to 
update models in each node. This method can compute the gradient accurately, but it requires heavy communication between nodes.
Also, it is shown that increasing minibatch size does not always benefit model training\cite{seide2011conversational}, 
especially in early stage of model training.

@@

\section{Model Averaging Strategy}
\subsection{All-reduce}
All-reduce strategy collects the weights from all the nodes. The communication is bounded by 
the bandwidth. Fig~\ref{fig:allreduce} is an example of all-reduce with 4 nodes. One node is responsible for
collecting all the model parameters. It will average the weights and then broadcast it to all the nodes. 
This approach compute the average weight based on all the information in computing nodes. Thus, we expect 
the converge speed to be the fastest.
\begin{figure}[htb]
  \centering
  \includegraphics[width=0.5\textwidth]{allreduce.png}
  \caption{All-reduce network}
  \label{fig:allreduce}
\end{figure}

This operation could be easily implemented by MPI\_Allreduce

@@

\section{Experimental Results}
\subsection{Setup}
In this paper, we report speech recognition results on the 300 hour Switchboard conversational telephone speech task 
(Switchboard-1 Release 2). We use MSU-ISIP release of the Switchboard segmentations and transcriptions (date 11/26/02),
together with the Mississippi State transcripts2 and the 30Kword lexicon released with those transcripts. 
The lexicon contains pronunciations for all words and word fragments in the training data. We use the Hub5 ’00 data for
evaluation. Specifically, we use the  the development set and Hub5 ’01 (LDC2002S13) data as a separate test set.

The Kaldi toolkit\cite{kaldi11} is used for speech recognition framework. Standard 13-dim PLP feature,
together with 3-dim Kaldi pitch feature, is extracted and used for maximum
likelihood GMM model training. Features are then transformed using LDA+MLLT before SAT training.
After GMM training is done, a tanh-neuron DNN-HMM hybrid system is trained using the the 40-dimension 
transformed fMLLR (also known as CMLLR \cite{gales1996generation}) feature as input and GMM-aligned senones 
as targets. fMLLR is estimated in an EM fashion for both training data and test data. A trigram language model (LM) is trained 
on 3M words of the training transcripts only.

Work in this paper is built on top of the Kaldi nnet1 setup and the NG-SGD method introduced in nnet2 setup. 
Details of DNN training follows Section 2.2 in \cite{vesely2013sequence}. In this paper, we use 6 hidden layers, where each 
hidden layer has 2048 neurons with sigmoids. Input layer is 440 dimension (i.e. the context of 11 fMLLR frames), 
and output layer is 8806 dimension. Mini-batch SGD is used for backpropagation and the minibatch is set to 1024 for all
the experiments. By defult, DNNs are initialized with stacked restricted Boltzmann machines (RBMs) that are pretrained 
in a greedy layerwise fashion \cite{hinton2006fast}. Comparison between random initialization and RBM-initialization 
in model averaging framework is reported in Section~\ref{sec:init}.

The server hardware used in this work is Stampede (TACC) (URL: https://portal.xsede.org/tacc-stampede). It is a Dell Linux 
cluster provided as an Extreme Science Engineering Discovery Environment (XSEDE) digital service by the Texas Advanced 
Computing Center (TACC). Stampede is configured with 6,400 Dell DCS Zeus compute nodes, the majority of which are configured
with two 2.7 GHz E5-2680 Intel Xeon (Sandy Bridge) processors and one Intel Xeon Phi SE10P coprocessor. 128 of the nodes are 
augmented with an NVIDIA K20 GPU and 8 GB of GDDR5 memory each, which we use for neural network training in this work.
Stampede nodes run Linux 2.6.32 with batch services managed by the Simple Linux Utility for Resource Management (SLURM).

\subsection{Averaging frequency}
Due to the limitation of computing resource, we did not tune averaging frequency in our setup. It is set to 1 averaging
per 20 minibatches by default. We would expect a tradeoff between lower averaging frequency (i.e. better speedup) and
better training accuracy. But the optimal point is yet to be explored.

\subsection{Learing Rate Schedule}
Initial learning rate is increased in porportion to number of threads in this setup. The reason for this is straight forward:
Assume we have $n$ minibatches of data for model training. When the model is trained using single thread, it gets updated $n$
times. When data is distributed to $m$ machines, then each model gets updated $n/m$ times. Since the effect of model averaging 
is mostly aggregating knowledge learnt from different data partition, the absolute change of model shall be compensated by 
$m$ times.

We compare two learning rate schedules in this section. The first one is the default setup used in Kaldi nnet1 (Newbob). 
It starts with a initial learning rate of 0.32 and halves the rate when the improvement in frame accuracy on a cross-validation 
set between two successive epochs falls below 0.5\%. The optimization terminates when the frame accuracy increases by less 
than 0.1\%. Cross-validation is done on 10\% of the utterances that are held out from the training data.

The second learing rate schedule is exponentially decaying. This method is used in \cite{senior2013empirical,povey2014parallel} 
and is shown to be superior to performance scheduling and power scheduling. In this work, it starts with the same initial learning
rate as the first method (Newbob), and decrease to the final learning rate (which is set to be 0.01 * initial learning rate). The
number of epochs is set to $15$ in this task, which is set to be the same as Newbob scheduling.

\begin{table}
  \centering
  \begin{tabular}{c|c|c|c|c}
    \hline
    & \multicolumn{2}{c|}{SWB}  & \multicolumn{2}{c}{CallHome} \\
    \hline
    Nodes       &  1      & 16    &  1    & 16 \\
    \hline
    Newbob      & 14.9    & 15.4  & 26.6  & 27.2 \\
    \hline
    exponential & 14.7    & 15.1  & 26.8  & 27.0 \\
    \hline
  \end{tabular}
  \caption{Comparing learning rate schedule}
  \label{tab:init}
\end{table}

\subsection{Initialization Matters}
\label{sec:init}
Compare random init and dbn init, single and parallell.

\begin{table}
  \centering
  \begin{tabular}{c|c|c|c|c}
    \hline
    & \multicolumn{2}{c|}{SWB}  & \multicolumn{2}{c}{CallHome} \\
    \hline
    Nodes          &  1     & 32     &   1    & 32\\
    \hline
    random init     & RR    & 16.4   &  RR    & 28.8 \\
    \hline
    rbm pretraining & 14.7  & 15.2   & 26.8   & 27.1 \\
    \hline
  \end{tabular}
  \caption{Comparing rbm pretraining with random initialization}
  \label{tab:init}
\end{table}

\subsection{Online NG-SGD Matters}
\begin{table}
  \centering
  \begin{tabular}{c|c|c|c|c}
    \hline
    & \multicolumn{2}{c|}{SWB}  & \multicolumn{2}{c}{CallHome} \\
    \hline
     Nodes  &  1    & 16        &   1   &  16 \\
    \hline
    SGD     & 14.9    & 16.3    &  26.9 & 28.3 \\
    \hline
    NG-SGD  & 14.7    & 15.1    &  26.8 & 27.0  \\
    \hline
  \end{tabular}
  \caption{Comparing NG-SGD and naive SGD}
  \label{tab:ngsgd}
\end{table}

@@

\begin{table}
  \centering
  \begin{tabular}{c|c|c|c|c|c|c}
    \hline
     Nodes  & 1    & 2    & 4    & 8    & 16 & 32\\
    \hline
Allreduce &    14.7 & -- & -- & 15.1 & 15.1   & 15.2\\
    \hline
Butterfly &    14.7 & -- & -- & --   & 15.6   & 16.0\\
    \hline
  \end{tabular}
  \caption{Comparison of WERs using different number of GPUs }
  \label{tab:wer}
\end{table}

It is shown that this is

@@

Further improvement might be achieved if parallel training runs on top of serial training initialization.
CUDA aware MPI, learning rate schedule. Averaging frequency. Theory on convergence.

@@

We acknowledge the Texas Advanced Computing Center (TACC) at The University of Texas at Austin for providing 
HPC resources that have contributed to the research results reported within this paper (URL: http://www.tacc.utexas.edu).
